---
title: "Lab 9: Testing functions, and string basics"
subtitle: "Stat 133, Spring 2019"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(testthat)
```

> ### Learning Objectives
>
> - Introduction to the R package "testthat"
> - Write simple functions and their unit tests
> - Test your code
> - String manipulation
> - Base R functions for strings

-----

The purpose of this lab is two-fold: 

1) Getting started with writing tests for
your functions (you don't need to submit any files for this part).

2) Getting started with functions to do basic manipulation 
of strings (you do NEED to submit Rmd and html files for this part).

-----

## R package `"testthat"`

`"testthat"` is one of the packages in R that helps you write tests for your functions. First, remember to install the package:

```{r eval = FALSE}
# do not include this code in your Rmd
# execute this in your console
install.packages("testthat")
```

- `"testthat"` provides a testing framework for R that is easy to learn and use
- `"testthat"` has a hierarchical structure made up of:
    + expectations
    + tests
    + contexts
- A __context__ involves __tests__ formed by groups of __expectations__
- Each structure has associated functions:
    + `expect_that()` for expectations
    + `test_that()` for groups of tests
    + `context()` for contexts

```{r eval = FALSE}
# load testthat
library(testthat)
```


### List of common expectation functions

| Function                  | Description                             |
|---------------------------|-----------------------------------------|
| `expect_true(x)`          | expects that `x` is `TRUE`              |
| `expect_false(x)`         | expects that `x` is `FALSE`             |
| `expect_null(x)`          | expects that `x` is `NULL`              |
| `expect_type(x)`          | expects that `x` is of type `y`         |
| `expect_is(x, y)`         | expects that `x` is of class `y`        |
| `expect_length(x, y)`     | expects that `x` is of length `y`       |
| `expect_equal(x, y)`      | expects that `x` is equal to `y`        |
| `expect_equivalent(x, y)` | expects that `x` is equivalent to `y`   |
| `expect_identical(x, y)`  | expects that `x` is identical to `y`    |
| `expect_lt(x, y)`         | expects that `x` is less than `y`       |
| `expect_gt(x, y)`         | expects that `x` is greater than `y`    |
| `expect_lte(x, y)`        | expects that `x` is less than or equal to `y` |
| `expect_gte(x, y)`        | expects that `x` is greater than or equal `y` |
| `expect_named(x)`         | expects that `x` has names `y`          |
| `expect_matches(x, y)`    | expects that `x` matches `y` (regex)    |
| `expect_message(x, y)`    | expects that `x` gives message `y`      |
| `expect_warning(x, y)`    | expects that `x` gives warning `y`      |
| `expect_error(x, y)`      | expects that `x` throws error `y`       |

-----

# Part 1) Practice writing simple tests

- To start the practice, create a new directory, e.g. `lab09`
- `cd` to `lab09`
- Create an `R` script to write and document your functions: e.g. `functions.R`
- Create a separate `R` script `tests.R` to write the tests of your functions.
- Below is the filestructure for this part of the lab assignment

```
  lab09/
    functions.R
    tests.R
```


### Example with `stat_range()`

Let's start with a basic function `stat_range()` to compute the overall range
of a numeric vector. Create this function in the file `functions.R`.

```{r}
#' @title Range
#' @description computes the range of a numeric vector (i.e. max - min)
#' @param x a numeric vector
#' @return the range value (max - min)
stat_range <- function(x) {
  max(x) - min(x)
}
```



### Tests for `stat_range()`

To write the unit tests in `tests.R`, we are going to consider the following 
testing vectors:

- `x <- c(1, 2, 3, 4, 5)`
- `y <- c(1, 2, 3, 4, NA)`
- `z <- c(TRUE, FALSE, TRUE)`
- `w <- letters[1:5]`

The typical structure of the tests has the following form:

```{r eval = FALSE}
# load the source code of the functions to be tested
source("functions.R")

# context with one test that groups expectations
context("Test for range value") 

test_that("range works as expected", {
  x <- c(1, 2, 3, 4, 5)
  
  expect_equal(stat_range(x), 4)
  expect_length(stat_range(x), 1)
  expect_type(stat_range(x), 'double')
})
```

- use `context()` to describe what the test are about
- use `test_that()` to group expectations:
    + output equal to 4
    + output of length one
    + output of type `double`
- to run the tests from the R console, use the function `test_file()` by passing
the path of the file `tests.R`

```{r eval = FALSE}
# assuming that your working directory is "lab09/"
library(testthat)
test_file("tests.R")
```

You could actually include a code chunk in your Rmd with the code above.


### Your Turn

Write more groups of tests---`test_that()`---to test `stat_range()` with the
rest of the testing vectors `y`, `z`, `w`:

- Using `y`, write expectations for:
    + output is of length one
    + output is equal to `NA_real_`
- Using `z`, write expectations for:
    + output is of length one
    + output is of type `integer`
    + output is equal to `1L`
- Using `w`, write expectations for:
    + throws an error


Try writing the following functions and come up with unit tests:

- `stat_centers()`
    - _title:_ measures of center
    - _description:_ computes measures of center such as Median and Mean
    - _param:_ a numeric vector
    - _return:_ a numeric vector with median and mean
    - use `mean()` and `median()` to write `stat_centers()`

- `stat_spreads()`
    - _title:_ measures of spread
    - _description:_ computes measures of spread such as Range, IQR, Standard Deviation
    - _param:_ a numeric vector
    - _return:_ a numeric vector with range, iqr, and stdev
    - use `stat_range()`, `IQR()`, and `sd()` to write `stat_spreads()`


-----

# Part 2) Basics of String Manipulation*

- This part of the lab involves working with strings in R.
- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab09-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab09-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.


In this second part of the lab,  you will be using the row names of the data frame `USArrests` (this data comes already in R):

```{r usarrests}
head(USArrests)

states <- rownames(USArrests)
head(states)
```

Here are some functions that you may need to use in this lab:

- `nchar()`
- `tolower()`
- `toupper()`
- `casefold()`
- `paste()`
- `paste0()`
- `substr()`


### Number of characters

`nchar()` allows you to count the number of characters in a string. Use it 
on `states` to get the number of characters of each state:


### Case folding

There are 3 functions to do case-folding: `tolower()`, `toupper()`, and `casefold()`. Apply each function on `states` to see what happens


### Length of State Names

We just saw how to use `nchar()` to count the number of characters in each 
state name:

```{r num-chars}
# number of charcaters
num_chars <- nchar(states)
```

Use the vector `num_chars` to obtain a frequency table called `char_freqs`, 
and then plot the frequencies with a bar chart:


### Pasting strings

R provides the `paste()` function. This function allows you to __paste__ 
(i.e. append, concatenate) character vectors separated by a blank space:

```{r pumpkin}
paste('Pumpkin', 'Pie')
```

You can give it any number of vector inputs

```{r abcde}
paste('a', 'b', 'c', 'd', 'e')
```

You can change the separator with `sep`

```{r abcde-sep}
paste('a', 'b', 'c', 'd', 'e', sep = '-')
```

`paste()` is vectorized:

```{r vectorized}
paste('a', 1:5, sep = '.')
```

There's a special wrapper around `paste()` called `paste0()` which is 
equivalent to `paste(..., sep = "")`

```{r paste0}
# paste0() -vs- paste(..., sep = "")
paste0('Pumpkin', 'Pie')
paste('Pumpkin', 'Pie', sep = '')

# paste0() is also vectorized
paste0('a', 1:5)
```


### Your Turn*: `paste()`

Use `paste()` to form a new vector with the first five states 
and their number of characters like this:

`"Alabama = 7"  "Alaska = 6"  "Arizona = 7"  "Arkansas = 8"  "California = 10"`


Now use `paste()`'s argument `collapse = ''` to _collapse_ the first five 
states like this:

`"AlabamaAlaskaArizonaArkansasCalifornia"`


### Your Turn*: Permutations

Find the number of permutations that can be formed by using all the given 
letters in `ALABAMA`. For instance: `"ALAMABA"`, `"AMALABA"`, `"AAAABML"`.



### Substrings

R provieds the function `substr()` to extract substrings in a character vector:

```{r substr-berkeley}
# extract first 3 characters
substr('Berkeley', 1, 3)
```

Use `substr()` to shorten the state names using the first 3-letters:


Use `substr()` to shorten the state names using the last 3-letters:


How would you shorten the state names using the first letter and the last
3-letters? For instance:  `"Aama"` `"Aska"` `"Aona"` `"Asas"` etc.



### Challenge

We already obtained a frequency table `char_freqs` with the counts of state 
names by number of characters. You can use those frequencies to get those 
state-names with 4-characters or 10-characters:

```{r}
# 4-char states
states[num_chars == 4]
```

```{r}
# 10-char states
states[num_chars == 10]
```

You can use `paste()` to join the 4-character states in one single string 
(i.e. _collapsing_) like this---separated by a comma and space---:

```{r}
# collapse 4-char states
paste(states[num_chars == 4], collapse = ", ")
```

Write code (using a for-loop) to obtain a 
list `states_list` containing the collapsed names by number of characters. 
If the number of characters is an even number, then the state names should be 
in capital letters. Otherwise, they should be in lower case letters.

Each list element of `states_list` must be named with the number of characters, 
followed by a dash, followed by the word `chars`: e.g. `'4-chars'`, `'5-chars'`,
etc. In total, `states_list` should have the same length as `char_freqs`.

Here's what `states_list` should look like for the first three elements:

```
$`4-chars`
[1] "IOWA, OHIO, UTAH"

$`5-chars`
[1] "idaho, maine, texas"

$`6-chars`
[1] "ALASKA, HAWAII, KANSAS, NEVADA, OREGON"
```

-----

## Converting from Fahrenheit Degrees

Here are four functions that convert from Fahrenheit degrees to other temperature scales:

```{r}
to_celsius <- function(x = 1) {
  (x - 32) * (5/9)
}

to_kelvin <- function(x = 1) {
  (x + 459.67) * (5/9)
}

to_reaumur <- function(x = 1) {
  (x - 32) * (4/9)
}

to_rankine <- function(x = 1) {
  x + 459.67
}
```

We can use the previous functions to create a more general function `temp_convert()`:

```{r}
temp_convert <- function(x = 1, to = "celsius") {
  switch(to,
         "celsius" = to_celsius(x),
         "kelvin" = to_kelvin(x),
         "reaumur" = to_reaumur(x),
         "rankine" = to_rankine(x))
}

temp_convert(30, 'celsius')
```

`temp_convert()` works fine when the argument `to = 'celsius'`. But what happens 
if you try `temp_convert(30, 'Celsius')` or `temp_convert(30, 'CELSIUS')`?


### Your turn*: Converting Temperatures

Rewrite `temp_convert()` such that the argument `to` can be given 
in upper or lower case letters. For instance, the following three calls should 
be equivalent:

```{r, eval=FALSE}
temp_convert(30, 'celsius')
temp_convert(30, 'Celsius')
temp_convert(30, 'CELSIUS')
```


-----

### Your Turn*: Names of files

Imagine that you need to generate the names of 10 data `.csv` files.
All the files have the same prefix name but each of them has a different number: 
`file1.csv`, `file2.csv`, ... , `file10.csv`. 

How can you generate a character vector with these names in R? Come up with at 
least three different ways to get such a vector:
```{r}
# vector of file names

```

Now imagine that you need to rename the characters `file` into `dataset`. In 
other words, you want the vector of file names to look like this: 
`dataset1.csv`, `dataset2.csv`, ... , `dataset10.csv`. Take the previous vector 
of file names and rename its elements:

```{r}
# rename vector of file names

```


-----

## Using function `cat()`

Run the following code:

```{r, eval=FALSE}
# name of output file
outfile <- "output.txt"

# writing to 'outfile.txt'
cat("This is the first line", file = outfile)
# insert new line
cat("\n", file = outfile, append = TRUE)
cat("A 2nd line", file = "output.txt", append = TRUE)
# insert 2 new lines
cat("\n\n", file = outfile, append = TRUE)
cat("\nThe quick brown fox jumps over the lazy dog\n",
    file = outfile, append = TRUE)
```

After running the previous code, look for the file `output.txt` in your working 
directory and open it. One of the uses of `cat()` is to write contents to a 
text file. Note that the first call to `cat()` does not include the argument 
`append`. The rest of the calls do include `append = TRUE`.


### Your turn*: `cat()` 

Modify the script such that the content of `output.txt` looks 
like the _yaml_ header of an `.Rmd` file with your information:

```
---
title: "Some title"
author: "Your name"
date: "today's date"
output: html_document
---

This is the first line
A 2nd line


The quick brown fox jumps over the lazy dog
```


-----

## Your Turn*: Valid Color Names

The function `colors()` returns a vector with the names (in English) of 657 
colors available in R. Write a function `is_color()` to test if a given 
name---in English---is a valid R color. If the provided name is a valid R color, 
`is_color()` returns `TRUE`.  If the provided name is not a valid R color 
`is_color()` returns `FALSE`.

```{r, eval=FALSE}
# your is_color() function



# test it:
is_color('yellow')  # TRUE

is_color('blu')     # FALSE

is_color('turkuoise') # FALSE
```


## Plot with a valid color

Use `is_color()` to create the function `colplot()` that takes one argument 
`col` (the name of a color) to produce a simple scatter plot with random
numbers (e.g. use `runif()` or `rnorm()` to get point coordinates).

If `col` is a valid name---say "blue"---, the scatterplot should show a title 
`"Testing color blue"`. 

If the provided `col` is not a valid color name, e.g. "blu", then the function 
must stop, showing an __error message__ `"invalid color blu"`.

```{r eval = FALSE}
# this should plot
colplot('tomato')

# this stops with error message
colplot('tomate')
```

-----

